Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
remove comments
update introduction

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

@@

Our research addresses this problem from multiple angles. In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting a desired patient behavior in a context of motivational interviews.

@@

As can be seen in Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm. Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. The latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification. In our experiment, we also employed two especial variant of RNN models: long short-term memory (LSTM) and gated recurrent unit (GRU).

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}

\textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
The architecture of proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization.

@@

%\subsection*{\textit{Impact of hidden states}}
%While MC is a non-parametric method, HMM has an important parameter, the number of hidden states. A natural question is how does this parameter affect the performance of HMM-based prediction  method. Figure~\ref{fig:hidden-states} provides an answer to this question by illustrating the performance of the proposed method using HMMs of different order in terms of F-measure on different types of sequences. As follows from this Figure, in general, the number of states does not significantly impact the classification performance. All four models show their best performance when the number of states is between 4 and 8. 
%
%Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\
%
%\begin{figure}[!htb]
%    \centering
%    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
%    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
%    \label{fig:hidden-states}
%\end{figure}
%
%\begin{table}[!htb]
%\centering
%\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
%\label{tab:emission_matrix_s}
%  \begin{tabular}{|l|l|}
%  \hline
%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
%1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
%2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
%  \end{tabular}
%\end{table} 
%
%\begin{table}[!htb]
%\centering
%\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
%\label{tab:emission_matrix_u}
%  \begin{tabular}{|l|l|}
%  \hline
%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
%1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
%2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
%  \end{tabular}
%\end{table}

@@

\begin{table}[h]
\centering
\caption{Common patterns in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Pattern} \\ \hline      
successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 306 CQ-EF: Closed question, Elicit Feedback $\rightarrow $ 120 HUP-O: High Uptake, other \\
& $\rightarrow $ 313 R-CML+: Reflect, commitment language positive $\rightarrow $ [positive commitment] \\\hline
unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 305 EA: Emphasize Autonomy $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
  \end{tabular}
\end{table}

>>>>

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov chain and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

However, Markov chain cannot handle long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step. Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps. On the other hand, Recurrent Neural Networks have no such limitations. RNNs are capable to capture long-term dependencies with their unlimited memory. It was shown that RNNs are very effective in sequence modeling with temporal data.  Recently, two variant of RNNs are proposed: Long Short-Term Memory (LSTM)\cite{graves2013speech} and Gated Reccurent Unit (GRU)\cite{chung2014empirical}, which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten,lipton2015learning, choi2016doctor}.

@@

Our research addresses this problem from multiple angles. In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting a desired patient behavior in a context of motivational interviews.

@@

As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm. Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. However, the latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variant of RNN models in our experiemnt: long short-term memory (LSTM) and gated recurrent unit (GRU).

@@

\textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

@@

\begin{table}[h]
\centering
\caption{Common patterns in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Pattern} \\ \hline      
successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
  \end{tabular}
\end{table}
